Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-07-12 Manuscript/manuscript.Rmd
results edits

![**Cell-line specific boundary summaries.** (A) Barplots illustrating that the total number of TAD boundary regions decreases as Hi-C data resolution decreases. (B) Venn diagrams showing that TAD boundary regions exhibit cell-line specificity across resolution. (C) Boxplots indicating that the proportion of class imbalance remained severe irrespective of resolution.](Figures/bounds_and_pis.png)

@@

In order to identify the optimal combination of data resolution, predictor type, and re-sampling technique necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 4). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line ($L = \{GM12878, K562\}$). Within each cell line we evaluated 4 data resolutions ($R = \{10 kb, 25 kb, 50 kb, 100 kb\}$). We then considered 3 separate predictor types as the feature space ($P = \{OC, OP, Distance\}$). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling ($B = \{None, RUS, ROS, SMOTE\}$). Each combination of the above cell line specific parameters was implemented on chromosome specific sections of the linear genome (omitting chromosome 9). Letting $|X|$ denote the cardinality [??? TBD] of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. For each combination of inputs, we performed 10-fold cross-validation to tune for the number of features to consider at each node of the random forest classifier. Each optimally tuned model was then validated on the same separate testing set for unbiased comparisons. Performances were evaluated using MCCs and aggregated across each chromosome by averaging. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

@@

## MODEL PERFORMANCE DECREASED AS Hi-C DATA RESOLUTION DECREASED

@@

![**SMOTE re-sampling coupled with distance-type predictors yielded optimal MCCs across resolution for cell line specific models.** Performances of random forest classification models built on each of the 3 predictor types including (A-B) OC, (C-D) OP, and (E-F) Distance, and across both the GM12878 and K562 cell lines. Within each cell line, predictor type, and resolution, chromosome-specific training data was re-sampled using different techniques including no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and *Synthetic Minority Over Sampling TEchnique* (SMOTE; purple). Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/model_performances_v2.png)

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models [??? Never defined, what is this?], within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

>>>>

![**Cell-line specific boundary summaries.** (A) Barplots illustrating that the total number of TAD boundary regions decreases as Hi-C data resolution decreases. (B) Venn diagrams showing that TAD boundary regions exhibit cell-line specificity across resolution. (C) Boxplots indicating that the proportion of class imbalance remained severe irrespective of resolution.](Figures/bounds_and_pis_v2.png)

@@

In order to identify the optimal combination of data resolution, predictor type, and re-sampling technique necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 4). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line ($L = \{GM12878, K562\}$). Within each cell line we evaluated 5 data resolutions ($R = \{5 kb, 10 kb, 25 kb, 50 kb, 100 kb\}$). We then considered 3 separate predictor types as the feature space ($P = \{OC, OP, Distance\}$). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling ($B = \{None, RUS, ROS, SMOTE\}$). Each combination of the above cell line specific parameters was implemented on chromosome specific sections of the linear genome (omitting chromosome 9). Letting $|X|$ denote the cardinality [??? TBD] of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2520 ($2 \times 5 \times 3 \times 4 \times 21$) separate models. For each combination of inputs, we performed 10-fold cross-validation to tune for the number of features to consider at each node of the random forest classifier. Each optimally tuned model was then validated on the same separate cell line specific testing set for unbiased comparisons. Performances were evaluated using MCCs and aggregated across each chromosome by averaging. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

@@

## MODEL PERFORMANCE PEAKED AT 10 KB AND DECREASED AS Hi-C DATA RESOLUTION DECREASED

@@

![**SMOTE re-sampling coupled with distance-type predictors yielded optimal MCCs across resolution for cell line specific models.** Performances of random forest classification models built on each of the 3 predictor types including (A-B) OC, (C-D) OP, and (E-F) Distance, and across both the GM12878 and K562 cell lines. Within each cell line, predictor type, and resolution, chromosome-specific training data was re-sampled using different techniques including no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and *Synthetic Minority Over Sampling TEchnique* (SMOTE; purple). Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/model_performances_v3.png)

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models [??? Never defined, what is this?; !!!defined above; should i mention again?], within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.
